---
title: "CGPA"
author: "Group 21"
date: "04/04/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

The objective of this study is to determine which socio-economic and demographic factors have
impact on students’ cumulative grade point average. 

# Libraries Used 
```{r}
library(tidyverse)
library(plyr)
library(dplyr)
library(readxl)
library(caret)
library(stats)
library(ggplot2)
library(ggpubr)
library(olsrr)
```

```{r}
setwd("~/Downloads")
gpa <- read_excel("GPAdata.xlsx", col_names = T)
gpa
summary(gpa)
```

# **Initial Data Analysis**

### **Distribution of numerical variables**

**Distribution of Credit**
```{r,message = FALSE, warning = FALSE}
ggplot(data = gpa) +
  geom_histogram(aes(x = Credit, fill = 'red'), col = 'black',show.legend = FALSE)

ggplot(data = gpa) + 
  geom_boxplot(aes(x = Credit), fill = 'skyblue' ,show.legend = FALSE)

```

The distribution of Credit is approximately symmetrical bell curve with very few outliers. There are about 4 outliers towards the right and 1 outlier at the left end of the distribution

**Distribution of CGPA**
```{r, message = FALSE, warning = FALSE}
ggplot(data = gpa) +
  geom_histogram(aes(x = CGPA, fill = 'red'), col = 'black', show.legend = FALSE)

ggplot(data = gpa) + 
  geom_boxplot(aes(x = CGPA), fill = 'skyblue' ,show.legend = FALSE)

```

The distribution of Credit is left skewed with very few outliers. There are about 4 outliers towards left end of the distribution.

###  **Distribution of categorical variables**

First, we define a function 'catplot()' that takes a categorical column and displays the bar plot of the distribution 

```{r, message = FALSE, warning = FALSE}
catplot = function(var){
 
ggplot(data = gpa) +
geom_bar(aes(x = get(var), y = (..count..)/sum(..count..), fill = 'red'), col = 'Black',show.legend = FALSE)+  xlab(as.character(var)) + ggtitle(paste0(" Bar plot of ", as.character(var)))+
theme_bw() + ylab("")
  
}

```

Run the same with different categorical variables.

```{r, message = FALSE, warning = FALSE}
p1 = catplot("Year")
p2 = catplot("Gender")
p3 = catplot("Status")
p4 = catplot("Major")
p5 = catplot("CommuteTime")
p6 = catplot("LiveFamily")
p7 = catplot("Interest")
p8 = catplot("SocialAct")
p9 = catplot("FamilyExp")
p10 = catplot("WeekWork")

gridExtra::grid.arrange(p1,p2,p3,p4, ncol = 2, nrow = 2)
gridExtra::grid.arrange(p5,p6,p7,p8, ncol = 2, nrow = 2)
gridExtra::grid.arrange(p9, p10, ncol = 1, nrow = 2)

```
# Correlation among variables
```{r}
#install.packages("corrplot")
#library(corrplot)
#head(gpa)
#gpa <- subset(gpa, select = -Year)
#M <- cor(gpa)
#corrplot(M, method = "circle", type = "upper")
```


```{r}
reg <- lm(CGPA ~ Credit + Year + Gender + Status + Major + CommuteTime + LiveFamily + Interest + SocialAct + FamilyExp + WeekWork, data = gpa)
anova(reg)
```

```{r}
colSums(is.na(gpa))
```


```{r}
gpa$Year <- factor(gpa$Year, order = TRUE,levels =c('3','4'))
gpa$Gender <- factor(gpa$Gender, levels = c('1','2','3'))
gpa$Status <- factor(gpa$Status, levels = c('0','1'))
gpa$Major <- factor(gpa$Major, levels = c('1','2','3'))
gpa$CommuteTime <- factor(gpa$CommuteTime, order = TRUE,levels = c('0','1','2','3','4','5'))
gpa$LiveFamily <- factor(gpa$LiveFamily, levels = c('0','1'))
gpa$Interest <- factor(gpa$Interest, order = TRUE,levels= c('1','2','3','4','5'))
gpa$SocialAct <- factor(gpa$SocialAct, order = TRUE, levels = c('1','2','3','4','5'))
gpa$FamilyExp <- factor(gpa$FamilyExp, order = TRUE, levels = c('1','2','3','4','5'))
gpa$WeekWork <- factor(gpa$WeekWork, order = TRUE)
gpa$WeekWork <- revalue(gpa$WeekWork,  c('15'="3",'18' = "4",'20'="4",'10'="2",'8' = "2",'30' = "5"))
```

```{r}
gpa %>% drop_na(Year) -> gpa
gpa %>% drop_na(Gender) -> gpa
gpa %>% drop_na(Status) -> gpa
colSums(is.na(gpa))
```

```{r}
classes <- sapply(gpa,class)
classes
```

```{r}
set.seed(1234)
gpa.sample <- sample(1:length(gpa$ID), 350, replace = FALSE)
train <- gpa[gpa.sample,]
test <- gpa[-gpa.sample,]
```

Model
```{r}
regR <- lm(CGPA ~ Credit + Year + Gender + Status + Major + CommuteTime + 
            LiveFamily + Interest + SocialAct + FamilyExp + 
            WeekWork,data = train)
anovaR <- anova(regR)
anovaR
```

```{r}
regF <- lm(CGPA ~ Credit + Year + Gender + Status + Major + CommuteTime + 
            LiveFamily + Interest + SocialAct + FamilyExp + WeekWork +
            Credit:Year + Credit:Gender + Credit:Status + Credit:Major +
            Credit:CommuteTime + Credit:LiveFamily + Credit:Interest + 
            Credit:SocialAct + Credit:FamilyExp + Credit:WeekWork, data = train)
anovaF <- anova(regF)
anovaF
```

Null hypothesis: coefficients of all interaction terms is equal to 0
Alternative hypothesis: at least one is not equal to 0
```{r}
SSEr <-anovaR$'Sum Sq'[12]
dfr <- anovaR$'Df'[12]
SSEf <- anovaF$'Sum Sq'[22]
dff <- anovaF$'Df'[22]
MSEf <-anovaF$'Mean Sq'[22]

fstat <- ((SSEr-SSEf)/(dfr-dff))/MSEf
pf(fstat, df1=dfr-dff, df2=dff, lower.tail = F)
#fail to reject null hypothesis, no significanct value to the interaction terms
# as part of the model
```

Final Model: CGPA ~ Credit + Major + CommuteTime + Interest + SocialAct
```{r}
library(MASS)
step <- stepAIC(regR, direction = "both"); step$anova
```
```{r}
final <- lm(CGPA ~ Credit + Major + CommuteTime + Interest + SocialAct, 
            data = train)
anova(final)
```

Validation MSE = 0.3093, MSPR = 0.29716, Since MSPR is approximately the same
as MSE then the final model is a valid selection to represent this 
relationship.
```{r}
pred <- predict(final, test[,c(3,7,8,10,11)])
delta <- gpa$CGPA[-gpa.sample]-pred
n.star <- dim(test)[1]
MSPR <- sum(delta^2)/n.star
MSPR
```

From the residual plot, you 
can see that the residuals do not look like a random scatter, they seem to be
more concentrated near the center of the graph, indicating potential non-constant variance.
The Normal Q-Q plot is not close to the line(y=x) near the top right most of the graph. 
So after conducting the Shapiro-Wilk test,
where the null hypothesis is that the residuals are normally distributed in some
population, the p-value =  0.01062 < 0.05. Thus, we reject the null-hypothesis and conclude that the residuals are 
not normally distributed in some population. In conclusion, this final model is not a good fit
to the data because the linearity assumption and non-constant variance may not be upheld.
```{r}
fit.resid <- final$residuals
fit.values <- final$fitted.values
plot(fit.values, fit.resid)
abline(0,0)

qqnorm(fit.resid)
qqline(fit.resid)

shapiro.test(fit.resid)
```


```{r}
library(MASS)
result = boxcox(final)
mylambda = result$x[which.max(result$y)]
mylambda

final2 <-lm(CGPA^(mylambda) ~ Credit + Major + CommuteTime + Interest + SocialAct, 
            data = train)
```

Boxcox transformation didn't work.
```{r}
fit.resid <- final2$residuals
fit.values <- final2$fitted.values
plot(fit.values, fit.resid)
abline(0,0)

qqnorm(fit.resid)
qqline(fit.resid)

shapiro.test(fit.resid)
```

```{r}
train$s.hat <- abs(final$residuals)
fit2 <- lm(s.hat~Credit + Major + CommuteTime + Interest + SocialAct, 
            data = train)
train$var.s <- (predict(fit2))^2

fit3 <- lm(CGPA~Credit + Major + CommuteTime + Interest + SocialAct, 
           weights = 1/var.s,
           data = train)
fit.resid <- fit3$residuals
fit.values <- fit3$fitted.values
plot(fit.values, fit.resid)
abline(0,0)

qqnorm(fit.resid)
qqline(fit.resid)

shapiro.test(fit.resid)
```

# Deleted Studentized Residual
```{r}
t <- rstudent(final)
alpha <- 0.05
n <- dim(train)[1]
p.prime <- length(coef(final))
t.crit <- qt(1-alpha/(2*n),n-p.prime-1)
round(t,2)
```

Since named integer(0) was returned after performing studentized deleted 
residuals test, therefore, none of these observations is an outlying CGPA 
observation.
```{r}
t.crit
which(abs(t) > t.crit)
```

```{r}
hii <- hatvalues(final)
round(hii,2)
```

```{r}
which(hii > 2*p.prime/n)
```

```{r}
which(hii > 0.5)
```

# Influential Observations
```{r}
DFFITS = dffits(final)
which(DFFITS >1)
```

```{r}
D = cooks.distance(final)
which(D > qf(0.2, p.prime, n-p.prime))
```

```{r}
DFFBETAS = dfbetas(final)
head(DFFBETAS)
```
```{r}
which(DFFBETAS > 1)
```

# Multicollinearity
Since there is no VIF value that exceeds 10, we can conclude that there is no indicative of serious multicollinearity
```{r}
library(car)
VIF = vif(final)
VIF
```
```{r}
VIFbar = mean(VIF)
VIFbar
```

# Lecture 21 slide 5
```{r}
t.rstandard <- t
t.rstudent <- rstudent(final)
t.inf <- influence.measures(final)
cbind(t.rstandard, t.rstudent)
```
# provides measures of detecting influential observation
```{r}
t.inf
```

# Graphical diagnostics
```{r}
i1 <- ols_plot_cooksd_chart(final) # cook's distance
```


```{r}
i2 <- ols_plot_resid_stud(final) # Deleted studentized residual
```


```{r}
i3 <- ols_plot_dfbetas(final) # DFbetas
```


```{r}
i4 <- ols_plot_dffits(final) # DIFFITS
```


```{r}
i5 <- ols_plot_resid_lev(final) # Studentized residual vs Leverages
```


```{r}
i6 <- ols_plot_resid_stud_fit(final) # Deleted studentized residual vs predicted value
```

```{r}
ggarrange(i1, i2, ncol = 2, nrow = 1)
```

```{r}
i3
```
```{r}
ggarrange(i4, i5, i6, ncol = 2, nrow = 2)
```

